# Chapter 5 -  Dimension reduction


```{r}
date()
```


## Data lookup

First loading the dataset and looking variable characteristics with summaries and histograms

```{r}
library(tidyr)
library(dplyr)
library(GGally)
library(corrplot)

human<-read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", header = T, sep = ",")

str(human)
head(human)
summary(human)

### Looking at histograms for variables
for (col in seq(from=1, ncol(human), by=3)) {
  par(mfrow = c(1,3))
  hist(human[,col], xlab = colnames(human)[col], main = colnames(human)[col] )
  hist(human[,col+1], xlab = colnames(human)[col+1], main = colnames(human)[col+1])
  if (col+2<= ncol(human)) 
    hist(human[,col+2], xlab = colnames(human)[col+2], main = colnames(human)[col+2])
}


```

The dataset is jointed from 2 datasets, human development "hd" and Gender inequality "gii", which info can be found here http://hdr.undp.org/en/content/human-development-index-hdi. The joined dataset contains 8 variables and 155 oservations, with countries as rownames. Here are short descriptions for variables:

Edu2.FM:      2nd female eductaion / 2nd Male education ratio
Labo.FM:      2nd female labor force participation / 2nd Male force participation ratio
Edu.Exp:      Expected years of schooling 
Life.Exp:     Life expectancy (years)
GNI:          Gross National Income 
Mat.Mor:      Eternal mortality rate
Ado.Birth:    Adolescent birth rate
Parli.F:      % of females in parliament

```{r}
ggpairs(human)
corrplot(cor(human),  method = 'square', order = 'FPC',  type = 'upper', diag = FALSE)
```

From the ggpairs and correlation plot we can see that there were several significant positive and negative correltions between variables.
Strong statistically significant positive correlations were observed between:
Adolescent births and maternal mortality, 
Life expectancy and 1) education expectancy, 2) higher female 2nd education ratio and 3) Gross National Income
Education expectancy and 1) higher female 2nd education ratio 2) Gross National Income
Higher female 2nd education ratio and Gross National Income

Strong statistically significant negative correlations were observed between:
Maternal mortality and 1) life expectancy, 2) education expectancy, 3) higher female 2nd education ratio, 4) Gross National Income
Adolescent births and all 4 variables that were negatively correlated with maternal mortality

## PCA on non-standardized data

Here, I'm performing Principal Component Analysis (PCA) with R's prcomp() default Singular Value Decomposition (SVD). THE PCA is an unsupervised dimension reduction method, in which data is first transformed to a new space with equal or less number of dimensions (PCs), and the 1st principal component captures the maximum amount of variance from the features in the original data, the 2nd principal component is orthogonal to the first and it captures the maximum amount of variability left, and so on. 
```{r}
# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))

```
Here PCA is visualized with biplot. Variation within the variables is shown with arrow length and correlation between the variables is shown with angle between arrows. As the data isn't normalized, most of the PC1/PC2 values are packed to upper right corner, with large outliers elsewhere.

## Standardizing the data, performing PCA and explaining PCs

Here I standardize the data with scale() function (mean of 0), and do another pca-biplot with variability explained by PCs shown on axis labels.  

```{r}
# standardize the variables
human_std <- scale(human)

# print out summaries of the standardized variables
summary(human_std)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_std)

# create and print out a summary of pca_human
s <- summary(pca_human)
s

# rounded percentages of variance captured by each PC
pca_pr <- round(100*s$importance[2,], digits = 1) 

# print out the percentages of variance
pca_pr

# create object pc_lab to be used as axis labels
pc_lab <- paste0(names(pca_pr), " (", pca_pr, "%)")

# draw a biplot
biplot(pca_human, cex = c(0.8, 1), col = c("grey40", "deeppink2"), xlab = pc_lab[1], ylab = pc_lab[2])

```
The PCs for standardized data look much better, as most of the observations are located at the middle, and it's much easier to spot the trends shown with the arrows.  

The 1st PC expalins 53.6% of the data variablilty, and 2nd PC explains 16.2% of the data. Combined, this is almost 70% of the variability in 8 variables, explained with just 2 (PCs)! 

Based on the plot, it looks like that most of variability in variables Mat.Mort and Ado.Birth (positive correlation and PC1 values) and Edu.Exp, Life.Exp, Edu.FM and GNI (positive correlation with each others, negative correlation with Mat.Mort and Ado.Birth and negative PC1 values) are explained in PC1
PC2 explains variablity in Labo.FM and Parli.Fm which have positive but PC2 vaues and positive, but weaker correlation with each other 

## Factominer's tea  data

Here we take a look at the "tea" data from Factominer-package. 

```{r}
library(FactoMineR)
data("tea")
str(tea)
head(tea)
```

the data contains 36 variables and 300 observations. Next, I select 6 columns for analysis.

```{r}
q# column names to keep in the dataset
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)
head(tea_time)

# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))

```
By looking at structure and few top observations, I can see that the selected variables are categorical, with 2-4 categories per variable.

Multiple Correspondence Analysis is another dimensionality reduction method. It's best suited for qualitative data with categorical variables, but continuous ones can be used as background (supplementary) variables. 

```{r}

# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali", graph.type = "classic")
```

  
Output of MCA summary contains (source: https://tuomonieminen.github.io/Helsinki-Open-Data-Science/#/69):  
- Eigenvalues with the variances and the percentage of variances retained by each dimension  
- The individuals coordinates, the individuals contribution (%) on the dimension and the cos2 (the squared correlations) on the dimensions.  
- Coordinates of the variable categories, the contribution (%), the cos2 (the squared correlations) and v.test value. The v.test follows normal distribution: if the value is below/above ± 1.96, the coordinate is significantly different from zero.  
- Categorical variables, which are squared correlation between each variable and the dimensions. If the value is close to one it indicates a strong link with the variable and dimension.

Summary shows that first 4 dimensions explain about haf of the variance.

From the MCA plot, I see that 29.47% of the variability are explained with Dim 1 and Dim 2. The plot shows correlations for variables, For example it seems that unpackaged tea (how) is almost always bought from tea shop (where) (bottom right), and the "how" and "where" variables appear together elsewhere in the plot, so they seem to have a strong correlation.